\section{The Colmena Execution Model}
\label{sec:model}

This section describes the machine on which the rest of the paper rests. The
production concerns we relocate in Sections~\ref{sec:context}--\ref{sec:hardening}
--- context lifecycle, credential handling, human-in-the-loop, and retry --- are
enforced not by convention but by a single execution engine, and it is the shape
of that engine, and of the agents it runs, that makes such enforcement possible.
We therefore introduce the execution model once, here, and later sections refer
back to it as the \emph{engine boundary} (\S\ref{sec:model}).

\subsection{An agent is a declarative document}
\label{subsec:declarative}

In Colmena an agent is not a program. It is a declarative JSON document that
describes a directed graph (a DAG augmented with optional bounded-retry
cycles) of typed \emph{nodes} connected by typed \emph{edges}. Nodes name
units of work; edges name the data that flows between
them. A node's \texttt{type} selects a fixed behavior implemented by the engine
rather than by the agent author: an \texttt{llm\_call} node performs a
model-driven tool loop; a \texttt{python\_script} node runs a fixed body of code;
a \texttt{secure\_suspend} node pauses execution to collect credentials and
returns them only as opaque handles; a \texttt{data\_run\_python} node executes
tabular code against an attachment inside a restricted sandbox; a \texttt{suspend}
node halts for a human decision; and a \texttt{log} node records an outcome. An
edge such as \texttt{draft.result $\rightarrow$ validate.draft\_json} routes one
node's named output to another node's named input; edges may be marked
\texttt{cyclic} to express bounded retry loops.

Listing~\ref{lst:secure-suspend} shows a real fragment from the secure-credential
agent used later in Section~\ref{sec:credentials}. The \texttt{secure\_suspend}
node declares which secrets to collect; nothing about how they are stored,
masked, or later injected appears in the document, because those are properties
of the node type, not of the agent.

\begin{lstlisting}[
  caption={A \texttt{secure\_suspend} node from \texttt{secrets\_agent.json}. The
    agent declares \emph{what} to collect; masking and injection are behaviors of
    the node type.},
  label={lst:secure-suspend},
  basicstyle=\ttfamily\small,
  breaklines=true,
  frame=single,
  captionpos=b]
"get_secrets": {
  "name": "get_secrets",
  "node_type": "secure_suspend",
  "node_schema": {
    "secrets": {
      "type": "array",
      "fixed": [
        { "name": "api_key",    "question": "Enter your API key" },
        { "name": "api_secret", "question": "Enter your API secret" }
      ]
    }
  }
}
\end{lstlisting}

Because the agent is described as data, its meaning is fully determined by the
node and edge types the engine understands; the author composes behaviors, but
does not implement the cross-cutting ones.

\subsection{One generic engine, and the boundary it creates}
\label{subsec:engine}

A single, generic Rust engine executes \emph{any} such document. Its public entry
point is \texttt{dag\_engine::\allowbreak api::\allowbreak run\_dag}, which takes
a DAG document and its
inputs, schedules nodes according to the edges, and drives each node to
completion. The engine contains no agent-specific logic: it is the same code path
whether it runs a refund workflow, a credential exchange, or a data-analysis
agent.

This uniformity is the paper's central mechanism. Because every agent is data
interpreted by one engine, that engine sits at a \emph{boundary} through which all
agent execution necessarily passes. A cross-cutting concern implemented at this
boundary --- how context is assembled and pruned, how a secret is masked before
it re-enters a model, when a human must approve, how a failed step is retried ---
holds for every agent by construction, because there is no execution path that
bypasses the engine. The agent author cannot forget to apply it, disable it, or
reimplement it inconsistently, since the behavior lives in the node type and the
engine, not in per-agent code. The invariants developed in
Sections~\ref{sec:context}--\ref{sec:hardening} are precisely properties made
enforceable by this boundary: each is a concern moved out of application code and
into the engine that interprets the DAG.

The engine core is written in Rust and is exposed to host languages through
foreign-function bindings --- a PyO3 binding surfaces \texttt{run\_dag} to Python
as \texttt{colmena.run\_dag}, and an napi binding surfaces it to
TypeScript/Node as \texttt{runDag} --- so the same engine, and the same boundary,
is the unit of execution regardless of the language that launches it.

\subsection{Deployment as a consequence of the model}
\label{subsec:deployment}

A practical consequence follows from treating agents as data. Adding a new agent,
altering an existing one, or rolling one back is, for the cross-cutting concerns
this paper addresses, a change to a configuration document rather than a code
deploy: the workflow is edited as a DAG and re-interpreted by the same engine,
with no change to the engine and no rebuild of its bindings. This does not mean an
agent carries no authored logic --- its prompts, routing rules, and any custom
tools remain engineered artifacts --- but the concerns we relocate to the engine
boundary are configured, not coded, and evolve with the document rather than with
a release of application code.
