\section{Context as an Engine-Managed Resource}
\label{sec:context}

The first concern we relocate to the engine boundary
(\S\ref{sec:model}) is the management of the model's context window. In a
multi-turn agent, the context window is a finite, shared resource that
accumulates the transcript of every prior turn: user messages, model
responses, and---decisively---the outputs of every tool the model has
invoked. Left unmanaged, that accumulation grows without bound, and each
subsequent turn re-transmits the entire history to the model, so the cost
of a session is quadratic in its length rather than linear. We call this
the \emph{context tax}. This section shows that Colmena discharges the tax
by engine default, and that a competitor framework can match the result
only by adding hand-rolled, per-tool application code.

\subsection{Design: context as a resource keyed off the graph}
\label{subsec:context-design}

Because every agent is a DAG interpreted by one engine
(\S\ref{sec:model}), the engine---not the agent author---decides how the
context window is assembled and pruned on each turn. Colmena treats the
window as a managed resource and applies, by default, a family of policies
keyed off the graph: attachments are passed by reference and materialized
only for the node that consumes them rather than retained inline;
binary tool outputs are scrubbed from the running history once produced;
the conversation history is compacted as it grows; and tool schemas are
loaded lazily rather than held resident for the whole session. None of
these behaviors is written into the agent document. They are properties of
the node types and of the engine that interprets them, and they therefore
hold for every agent that passes through the boundary, with no per-agent
code to author, apply, or forget. The remainder of this section isolates
the single policy that dominates the cost---the scrubbing of binary
tool outputs---and measures it.

\subsection{The default tax}
\label{subsec:context-default}

We evaluate the policies on a fixed ten-turn analyst session, run
identically across every framework through the measurement harness of
\S\ref{sec:methodology}. The session loads a document,
\texttt{Q3\_2026\_report.md} of $\approx$\num{3024} tokens across eleven
sections, and exposes a single tool, \texttt{generate\_chart}, that
returns a fixed $\approx$\num{8168}-token opaque base64 blob. The model is
driven to call the tool on three of the ten turns (turns three, six, and
nine). Every framework runs the same workload; only the engine's
context management differs.

Table~\ref{tab:context-totals} reports the total input tokens consumed
over the session. The competitor frameworks cluster between \num{404095}
and \num{452359} input tokens; Colmena consumes \num{39085}, an order of
magnitude fewer, at a session cost of \$0.018 against \$0.125--\$0.142 for
the competitors.

\begin{table}[t]
  \centering
  \caption{Total input tokens over the fixed ten-turn analyst session
    (\S\ref{sec:methodology}). The competitor frameworks re-transmit
    retained binary tool outputs on every subsequent turn; Colmena scrubs
    them by engine default. The Google ADK \texttt{artifacts\_scrub} arm
    adds $\approx$15--20 lines of per-tool application code that reproduces
    the effect (mean of $N=3$; \S\ref{subsec:context-closure}).}
  \label{tab:context-totals}
  \begin{tabular}{lS[table-format=6.0]}
    \toprule
    Framework & {Input tokens} \\
    \midrule
    Colmena (engine default)        & 39085 \\
    \midrule
    LangGraph                       & 404095 \\
    LlamaIndex                      & 419934 \\
    Google ADK                      & 445370 \\
    LangChain                       & 452158 \\
    CrewAI                          & 452359 \\
    \midrule
    Google ADK, \texttt{artifacts\_scrub} & 28795 \\
    \bottomrule
  \end{tabular}
\end{table}

The escalation is not caused by the document. The document is loaded once
and is small; a plain document turn costs $\approx$\num{3600} input tokens
before any chart is produced. The cost is localized to the retained binary
tool outputs. Tracing the LangChain arm turn by turn exposes the mechanism
as a staircase: a document turn costs $\approx$3.6k tokens before the first
chart, then $\approx$26k after chart one, $\approx$48k after chart two, and
$\approx$71k after chart three. Each chart adds its $\approx$\num{8168}-token
blob to the history, and because the full history is re-transmitted on every
following turn, each retained blob is paid for again on each subsequent
turn. The staircase, not the document, is the tax, and its steps coincide
exactly with the three chart turns.

\subsection{The closure}
\label{subsec:context-closure}

The gap in Table~\ref{tab:context-totals} is fully closable in a
competitor framework. We demonstrate this directly. In the Google ADK
\texttt{artifacts\_scrub} arm we hand-wrote a tool that calls
\texttt{save\_artifact} on the PNG and returns only a handle in place of
the blob, together with a \texttt{load\_artifacts} step for the document.
This is the same behavior Colmena's engine applies by default, but written
as per-tool application code---roughly fifteen to twenty lines. With it,
ADK consumes \num{27904}, \num{30011}, and \num{28470} input tokens across
three runs, a mean of \num{28795} ($N=3$), against the \num{445370} of its
default arm.

% TODO(Task 4): \ref{fig:context}
The mechanism is visible in the per-turn token curves of
Figure~\ref{fig:context}, which overlays three arms: Colmena's default,
ADK's default, and ADK with \texttt{artifacts\_scrub}. On the chart turns
the default ADK arm spikes---$\approx$19k, $\approx$75k, $\approx$120k
input tokens---as retained blobs compound, while both the Colmena and the
\texttt{artifacts\_scrub} curves stay flat, at $\approx$1.3--2.2k tokens on
those same turns. The token counts are the evidence: the proxy spans of
\S\ref{sec:methodology} are metadata only, and it is the measured token
curve, not any transcript text, that establishes where the cost is spent.
All three arms pass the task quality guardrail of \S\ref{sec:methodology}.

\subsection{Reading}
\label{subsec:context-reading}

This concern is the first instantiation of the paper's default-versus-code
thesis. Colmena delivers by engine default---with no line of agent
code---the same context economy that the ADK reaches only by hand-rolled,
per-tool application code. The \texttt{artifacts\_scrub} arm is not a
Colmena weakness but the evidence for the thesis: it shows the win is real
and reproducible, and it prices the alternative at fifteen to twenty lines
of application code that every tool, in every agent, must carry and keep
correct. What Colmena moves to the engine boundary, the competitor must
re-implement per tool, and remember to.
