\section{Credential Isolation}
\label{sec:credentials}

The second concern we relocate to the engine boundary
(\S\ref{sec:model}) is the isolation of secrets from the model context. A
production agent routinely handles credentials the user must supply at run
time---an API key, a one-time authorization token, a password---and then
forwards them to some downstream tool or endpoint. The hazard is
structural: in the idiomatic construction of every mainstream framework,
the credential is collected by, and returned through, the language model
itself, and so it is written into the transcript that every subsequent turn
re-transmits. A secret in the context window is a secret in the logs, in
the provider's request history, and in every cache the transcript touches.
This section shows that Colmena forecloses that exposure by engine default,
and that a competitor framework can match the guarantee only by
hand-architecting the credential path out of the model's reach.

\subsection{The invariant: two enforcement points}
\label{subsec:cred-invariant}

Colmena enforces a single invariant---\emph{no plaintext secret ever
enters the model context}---and it enforces it at the two points on the
credential's round trip where a secret would otherwise cross the boundary
into the transcript.

The first is the \emph{inbound} path. When a node requires a credential,
the engine's \texttt{secure\_suspend} primitive pauses execution, collects
the value out of band from the user, and routes it directly to the
consuming tool. The language model is never asked to gather the secret and
never sees it: the collected value is bound to the tool invocation by the
engine, not passed through a model turn. The model orchestrates
\emph{that} a credential is needed and \emph{where} it goes, but never
handles the value itself.

The second is the \emph{outbound} path. A tool's own result may itself
carry the secret back---an authentication endpoint that echoes the
submitted token in its success response is the canonical case. Here the
engine applies tool-result masking: designated fields, such as the
\texttt{auth\_token} of the response, are scrubbed from the result before
it re-enters the running history. The credential reaches the tool and the
tool reaches the endpoint, but the value never returns to the context.

One invariant, two enforcement points. Both are properties of the engine
that interprets the graph, not of application code the agent author writes,
and both therefore hold for every agent that passes through the boundary.
We consolidate the entire secret story here: the outbound masking is owned
by this section, and \S\ref{sec:hardening} only references it.

\subsection{Result: the leak matrix}
\label{subsec:cred-result}

We evaluate the invariant on a credential-collection task run identically
across every framework through the measurement harness of
\S\ref{sec:methodology}. The task presents two attack surfaces in one
workload: an inbound cell, in which the agent must collect a secret from
the user and deliver it to a downstream endpoint, and an outbound cell, in
which the endpoint echoes the submitted credential back in its response.
Each surface is run at three seeds, giving six cells per framework. The
criterion is binary and proxy-audited (\S\ref{sec:methodology}): a cell is
marked \emph{leaked} if the plaintext secret is observed anywhere in the
traffic to the model provider.

The comparison is fair by construction. The task is identical for every
framework, and every framework \emph{delivers} the credential to the
endpoint---delivery is $6/6$ across the board, including Colmena. Colmena
does not evade the leak by declining the task; it performs the same work as
the competitors and simply keeps the value off the model path while doing
it. Table~\ref{tab:cred-leak} reports the outcome.

\begin{table}[t]
  \centering
  \caption{Secret leakage and task delivery on the credential-collection
    task (\S\ref{sec:methodology}), $N=3$ seeds $\times$ two attack surfaces
    $=6$ cells per framework. A cell is \emph{leaked} if the plaintext
    secret appears in the proxy-audited traffic to the model provider.
    Every framework delivers the credential to the endpoint; the five
    idiomatic-default competitors leak it on every cell. The
    \texttt{langgraph\_interrupt\_isolated} arm is a hand-architected
    steelman (\S\ref{subsec:cred-steelman}).}
  \label{tab:cred-leak}
  \begin{tabular}{lS[table-format=1.0]S[table-format=1.0]}
    \toprule
    Framework & {Leaked} & {Delivered} \\
    \midrule
    Colmena (engine default)                  & 0 & 6 \\
    \midrule
    CrewAI                                    & 6 & 6 \\
    Google ADK                                & 6 & 6 \\
    LangChain                                 & 6 & 6 \\
    LangGraph                                 & 6 & 6 \\
    LlamaIndex                                & 6 & 6 \\
    \midrule
    LangGraph, \texttt{interrupt\_isolated}   & 0 & 6 \\
    \bottomrule
  \end{tabular}
\end{table}

The separation is total. Colmena leaks the secret on $0$ of $6$ cells; each
of the five competitor frameworks, in its idiomatic default construction,
leaks on all $6$. This is not a matter of degree or of tuning: the default
credential path in every competitor runs the secret through the model, so
every seed on both attack surfaces leaks, while the engine invariant admits
no cell in which the secret crosses the boundary.

\subsection{Honest steelman: the invariant is reachable by construction}
\label{subsec:cred-steelman}

The $100\%$ leak rate is a property of the \emph{default} construction, not
a hard limit of the competitor frameworks. We demonstrate this directly.
The \texttt{langgraph\_interrupt\_isolated} arm reaches $0/6$, matching
Colmena exactly, while still delivering $6/6$. It does so by
hand-architecting the credential path out of the model's reach: it uses
LangGraph's \texttt{interrupt()} primitive to collect the secret out of
band and a checkpointer to carry it across the pause, so the value is bound
to the downstream call without ever passing through a model turn. This is
the same mechanism the Colmena engine applies by default, rebuilt as
explicit graph-construction code.

This arm is not a Colmena weakness but the evidence for the paper's
default-versus-code thesis, in the same form it took for context
(\S\ref{sec:context}). A competitor \emph{can} be safe---the invariant is
reachable---but only by an author who knows the hazard, reaches for the
out-of-band primitive, and wires the isolated path by hand, for this
credential and every other. What Colmena guarantees at the engine boundary
for every agent that crosses it, the competitor must architect per agent,
and remember to. The safe outcome is a default on one side of the boundary
and a construction task on the other.
