\section{Production Hardening as Declarative Configuration}
\label{sec:hardening}

\newcommand{\cmark}{\ensuremath{\bullet}}

The third and final concern we relocate to the engine boundary
(\S\ref{sec:model}) is the collection of capabilities that separate a
demonstration agent from a production one: a durable human approval gate,
an automatic critic-retry loop, and the isolation of secrets from the model
context. Unlike the first two concerns, this one is not a performance or a
leakage gap. Every framework we measure can be made to exhibit every one of
these capabilities, and in our experiment all of them do. The distinction
this section develops is therefore not one of \emph{whether} the capability
is reachable but of \emph{how} it is authored and \emph{what} that authoring
guarantees: in Colmena each capability is a property of the graph structure
the engine interprets, whereas in every competitor it is imperative
control-flow code the agent team writes, tests, and maintains by hand.

\subsection{Design: hardening as graph structure}
\label{subsec:hardening-design}

Colmena expresses each production capability as a feature of the declarative
agent document rather than as application logic. A durable human-in-the-loop
approval gate is a suspend node: the engine's \texttt{secure\_suspend}
primitive pauses execution at the node, persists the run state, and resumes
it---across a fresh process, if need be---when the out-of-band decision
arrives, so the pause survives the lifetime of any single worker. A
critic-retry gate is a cyclic edge: the graph routes a candidate answer
through a critic node and, on rejection, back to the producing node, so the
retry loop is drawn in the graph rather than coded as a bounded \texttt{while}
around a model call. Credential masking is a flag on the secure path,
enforced at the engine boundary; its mechanism and its measured result are
owned by \S\ref{sec:credentials}, and we do not re-argue them here. In each
case the behavior lives in the node and edge types the engine understands
(\S\ref{sec:model}), not in per-agent code.

The competitor frameworks reach the same three capabilities by
hand-rolling them as imperative control flow. The approval gate becomes an
explicit interrupt-and-checkpoint construction the author wires around the
model turn; the critic-retry loop becomes a hand-written control structure
that re-invokes the producer on a failed check; the credential path becomes
the out-of-band routing \S\ref{sec:credentials} anatomizes. This code is
entirely writable---and, as the evaluation below shows, it works---but it is
code, authored once per agent and owned thereafter by the team that wrote it.

\subsection{Evaluation (EQ3): capability parity across frameworks}
\label{subsec:hardening-eval}

We evaluate the three capabilities on a refund-authorization task run
identically across every framework through the measurement harness of
\S\ref{sec:methodology}. A customer requests a \$250 refund for a
duplicate charge against a policy that auto-approves at most \$100
per agent; the correct disposition is therefore to \emph{escalate} rather
than to approve. Each framework must reach that decision, route the
escalation through a durable human approval gate that survives a
process boundary, pass its candidate answer through a critic gate, and
keep the collected credential off the model path. We score four binary
criteria per framework---\emph{decision} (the escalate disposition is
correct under policy), \emph{HITL} (the suspend/resume gate completes
across a fresh process), \emph{critic} (the critic-retry gate fires), and
\emph{masking} (no plaintext secret reaches the model provider, audited as
in \S\ref{sec:credentials})---together with their conjunction
\emph{all\,ok}.

Every framework achieves every capability. Table~\ref{tab:hardening-caps}
reports the result: all six frameworks reach the correct escalate decision,
complete the durable approval gate, fire the critic gate, and mask the
secret, so \emph{all\,ok} holds for each. This is the point of the
experiment, not a limitation of it. The production capabilities are not a
capability gap between Colmena and the field; they are reachable, and
reached, everywhere. What differs is the authoring model behind identical
outcomes.

\begin{table}[t]
  \centering
  \caption{Capability results on the refund-authorization task
    (\S\ref{sec:methodology}), scored as four binary criteria per framework
    plus their conjunction. \emph{Dec.}: the escalate disposition is correct
    under the \$100 policy. \emph{HITL}: the suspend/resume approval
    gate completes across a fresh process. \emph{Critic}: the critic-retry
    gate fires. \emph{Mask}: no plaintext secret reaches the model provider
    (\S\ref{sec:credentials}). Every framework achieves every capability
    (\cmark\ = pass).}
  \label{tab:hardening-caps}
  \begin{tabular}{lccccc}
    \toprule
    Framework & Dec. & HITL & Critic & Mask & All\,ok \\
    \midrule
    Colmena     & \cmark & \cmark & \cmark & \cmark & \cmark \\
    \midrule
    CrewAI      & \cmark & \cmark & \cmark & \cmark & \cmark \\
    Google ADK  & \cmark & \cmark & \cmark & \cmark & \cmark \\
    LangChain   & \cmark & \cmark & \cmark & \cmark & \cmark \\
    LangGraph   & \cmark & \cmark & \cmark & \cmark & \cmark \\
    LlamaIndex  & \cmark & \cmark & \cmark & \cmark & \cmark \\
    \bottomrule
  \end{tabular}
\end{table}

The authoring model is what the code-volume comparison exposes, and it must
be read precisely. Table~\ref{tab:hardening-loc} splits, for each framework,
the lines of imperative code from the lines of declarative configuration the
solution comprises. Colmena's solution is \emph{not} the smallest: it
authors \num{120} lines of code and \num{115} lines of configuration,
\num{235} lines in total, more than every competitor's single-column count.
We claim no line-count advantage, and the reader should draw none. The
comparison is qualitative, not quantitative: the competitors carry their
hardening entirely as imperative code---\numrange{93}{171} lines of it, and
\emph{zero} lines of configuration---whereas the bulk of Colmena's logic,
\num{115} lines, is declarative configuration that the generic engine
interprets rather than control flow the team runs. The two columns measure
different kinds of authoring, and it is the kind, not the size, that this
section is about.

\begin{table}[t]
  \centering
  \caption{Authoring-model split for the refund-authorization solution:
    lines of imperative code versus lines of declarative configuration per
    framework. Colmena is not the smallest solution---it authors the most
    lines in total---but the majority of its logic is configuration the
    engine interprets, whereas every competitor carries its hardening as
    imperative code with no configuration. The comparison is of the kind of
    authoring, not of size.}
  \label{tab:hardening-loc}
  \begin{tabular}{lS[table-format=3.0]S[table-format=3.0]}
    \toprule
    Framework & {Code (LOC)} & {Config (LOC)} \\
    \midrule
    Colmena     & 120 & 115 \\
    \midrule
    CrewAI      &  93 &   0 \\
    LangChain   &  99 &   0 \\
    LlamaIndex  &  99 &   0 \\
    Google ADK  & 117 &   0 \\
    LangGraph   & 171 &   0 \\
    \bottomrule
  \end{tabular}
\end{table}

\subsection{The guarantee, and its deployment consequence}
\label{subsec:hardening-guarantee}

Capability parity is genuine, but it is parity of \emph{outcome on a passing
run}, and the two authoring models differ in what they guarantee off that
run. Colmena's approval gate, critic loop, and masking are invariants of the
engine that interprets the graph (\S\ref{sec:model}): they are keyed off the
declarative structure and hold for every agent that crosses the boundary,
with no execution path that bypasses them. The competitors' equivalents are
hand-rolled code that each team must write, test, and maintain per agent, and
hand-rolled safety code is omittable and can drift. The credential
counterfactual of \S\ref{sec:credentials} is the direct proof: there the
competitors' \emph{default} construction leaks the secret on every trial, and
the safe outcome is recovered only by an author who knows the hazard and wires
the isolated path by hand---the same capability, present when coded and absent
when not. A passing row in Table~\ref{tab:hardening-caps} records that the
code was written correctly for this agent, on this run; it does not, as the
engine invariant does, guarantee the capability for the next agent the team
authors.

The deployment consequence follows from where the logic lives. Because
Colmena's hardening is configuration the engine interprets rather than code
the application runs, changing one of these cross-cutting behaviors---
tightening the approval policy, adjusting the critic gate, extending the
masked path---is an edit to the agent document, re-interpreted by the same
engine, rather than a code deploy (\S\ref{subsec:deployment}). This is
scoped precisely to the cross-cutting concerns this paper relocates: an
agent still carries authored prompts, routing rules, and custom tools, and
those remain engineered artifacts. What moves to the configuration surface
is the production hardening---and with it, the guarantee that the hardening
holds is moved off the shoulders of each agent's authors and onto the engine
boundary they all share.
