\section{Limitations and Negative Results}
\label{sec:limitations}

The preceding sections argue that Colmena relocates three production
concerns to engine-enforced invariants and measure the resulting gains.
Honest evaluation requires reporting where the comparison does not favor
Colmena and stating plainly what the engine's guarantees do, and do not,
establish about the competitors. Two results follow.

\subsection{Speed: round-trips traded for token volume}
\label{subsec:limitations-speed}

Colmena's per-call overhead is not the source of its token advantage. A
single-tool-call calibration task ("hello world", ten trials per
framework) isolates fixed per-call cost from the multi-turn dynamics of
\S\ref{sec:context}: median per-call latency is at parity between Colmena
and the fastest competitor---\SI{761}{ms} for Colmena versus \SI{738}{ms}
for LangChain. Colmena's engine is not slower to answer a single call.

What differs is the number of calls a session makes. Over the ten-turn
context-tax session of \S\ref{sec:context}, Colmena issues approximately
$18$ model round-trips against approximately $13$ for the competitor
frameworks. The extra calls are a direct consequence of the same
mechanisms that produce Colmena's token economy: lazy tool loading turns
each on-demand schema or attachment fetch into its own model turn rather
than folding it into an existing one, and history compaction periodically
interposes a summarization turn in place of re-sending the full transcript.
Both mechanisms trade a round-trip for the tokens that round-trip avoids
resending on every subsequent turn.

The net effect is that Colmena is \emph{not} faster per session. Holding
per-call latency at parity, more calls means more wall-clock time and more
round-trip overhead than a framework that resends a larger context in
fewer, larger calls. This is a genuine trade-off, not an artifact of
measurement: the same lazy loading and compaction that yield the
order-of-magnitude token reduction of \S\ref{sec:context} are what add the
extra round-trips. We report the exchange rate rather than obscure it---
Colmena buys token volume with call count, and a deployment that is
latency-bound on session wall-clock time rather than cost-bound on token
volume should weigh the trade in that direction.

\subsection{The central caveat: enforcement, not exclusivity}
\label{subsec:limitations-caveat}

Every engine-enforced invariant this paper measures is, on inspection,
achievable by a competitor framework through hand-written code. We do not
merely assert this; each of the two preceding results sections
demonstrates it directly. The Google ADK \texttt{artifacts\_scrub} arm of
\S\ref{sec:context} hand-rolls a per-tool artifact handle in place of the
raw binary output and closes the context-tax gap against Colmena's
default. The \texttt{langgraph\_interrupt\_isolated} arm of
\S\ref{sec:credentials} hand-architects the credential path out of the
model's reach using LangGraph's \texttt{interrupt()} primitive and a
checkpointer, and reaches the same $0\%$ leak rate as Colmena's default.
In both cases a competitor framework, correctly used by an author who
anticipates the hazard and writes the mitigating code, matches the
outcome the Colmena engine guarantees without any such code.

The contribution this paper measures is therefore not that competitor
frameworks are \emph{incapable} of the safe or efficient outcome. It is
that Colmena makes that outcome the \emph{default}, enforced once at the
engine boundary (\S\ref{sec:model}) for every agent that crosses it,
whereas every competitor reaches the same outcome only as a
\emph{construction}---code that a specific author must know to write,
must test, and must maintain, separately, for every tool and every agent
that needs it. The steelman arms are not exceptions that weaken the
paper's measurements; they are the evidence that isolates what the
measurements attribute to the engine. What Colmena guarantees by default,
the field can reach by hand, and the gap this paper reports throughout is
the gap between an invariant an author cannot omit and a practice an
author must remember, and remember correctly, every time.
