\section{Conclusion}
\label{sec:conclusion}

Production agents accrue concerns that prototypes do not: a context that
grows without bound, credentials that pass through the model on their way to
a tool, and the durability, supervision, and self-correction that separate a
demonstration from a service. Mainstream frameworks leave each of these to
per-application code. This paper has argued that Colmena instead relocates
them into invariants that a single execution engine enforces by
construction, and that the declarative DAG-as-data agent is what makes such
relocation possible: because the agent is data the engine interprets rather
than code it merely hosts, the engine can guarantee a property of every
agent it runs rather than trusting each application to supply it.

The three concerns we measured instantiate one thesis rather than three
separate findings. Context becomes an engine-managed resource, and the safe
size is the one the engine keeps by default. Credential isolation becomes a
property of a node type, and the transcript that never sees a plaintext
secret is the one the engine produces by default. Production hardening
becomes a feature of the graph the engine interprets, and the durable,
supervised, self-correcting agent is the one the configuration already
describes. In each case the concern moves from something an author must
remember to write, test, and maintain into something that holds without the
author's attention---so that the safe and efficient path is not a discipline
imposed on the agent but the default the engine follows. Our own honest
accounting sharpens rather than softens this claim: each outcome is
reachable by a competitor framework in hand-written code, and what the
engine contributes is precisely that no author has to write it.

We offer this as a design stance for agent frameworks. Where a framework can
express an agent as data it interprets rather than as code it hosts, the
production concerns that today are left to convention can instead be made
invariants of the execution model---enforced once, uniformly, for every
agent that crosses the engine boundary.
